\documentclass{article}
\usepackage[top=1in,bottom=1in]{geometry}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\graphicspath{ {./assets/} }
\usepackage[none]{hyphenat}
\date{}
\title{MUSC360 Music in Western Culture Before 1900: Music as a Public Art}
\begin{document} 
  \author{Michael Li}
  \title{MUSC360 Music in Western Culture Before 1900: Music as a Public Art}
  \maketitle
  \tableofcontents
  \newpage
  \section{Christimas in Paris, ca. 1280}
  During this time, Paris was one of Europe's largest cities and was a cultural and educational center.
  \begin{itemize}
    \item University of Paris (provided liberal arts education for white collar jobs)
    \item Cathedral of Notre Dame
  \end{itemize}
  The university and cathedral grew together with the literature culture, promoting poetry and debate. In particular, music was everywhere in Paris, especially amongst the aristocrats and church.
  \subsection{Church Music}
  Church music was performed in 2 main settings
  \begin{itemize}
    \item \textbf{The Office}: set of services in church conducted by and for the clergy (no laymen allowed).
    \item \textbf{The Mass}: daily service in church performed by professionals for laymen. Works were performed in Latin and included
      \begin{itemize}
        \item Spoken texts (e.g. readings and sermons)
        \item Sung texts (e.g. prayers and psalmody) set to \textbf{Gregorian chant}
      \end{itemize}
      Chants in particular can be broken down into 2 types:
      \begin{itemize}
        \item \textbf{Action chants} that accompany actions and rituals
        \item \textbf{Lesson chants} performed after readings to foster reflection of learned material.
      \end{itemize}
      Mass texts can be be broken down into 2 types:
      \begin{itemize}
        \item \textbf{Mass Ordinary}: texts performed every day
        \item \textbf{Mass Proper}: texts chosen according to the day of the liturgical calendar, usually from the \textbf{Book of Psalms} which has proper chants called \textbf{psalmody}.
      \end{itemize}
      Gregorian chants today are compiled in \textbf{Liber Usualis} and are notated using \textbf{square notation}.
  \end{itemize}
  \subsection{Puer natus est nobis (“A child is born to us”)}
  First chant of Christmas Day Mass. It is an \textbf{Introit}, which is an action chant\\
  It is proper and corresponding texts from Psalm 97 line 1 and the Book of Isaiah \\
  Performed as an \textbf{Antiphonal psalmody} (performed by 2 halves of the choir in alternation)
  \begin{itemize}
    \item \textbf{monophonic} in texture (only 1 line sung in unison)
    \item Pitch organization: \textbf{mode 7} (final - G, reciting tone - D). Reciting tone is the note that is repeated a lot.
    \item Melody is constructed from 
      \begin{itemize}
        \item Chanting on reciting tone
        \item \textbf{Melodic formulas} (melodic material shaped around punctuation and the grammar of the text).
      \end{itemize}
    \item Rhythmic organization is unknown.
  \end{itemize}
  Structure consists of
  \begin{itemize}
    \item Antiphon (opening segment that is based on \textbf{neumatic} text and multiple notes are sung on 1 syllable)
    \item Verse (marked Ps. and based on \textbf{syllabic} text where multiple syllables are sung on one note)
    \item Lesser Doxology (short repeated text that is required in every chant that they do, in the same tune as the verse and is \textbf{syllabic})
    \item Repeat of Antiphon (2nd choir marked by *)
  \end{itemize}
  \subsection{Viderunt omnes (“All the ends of the earth have seen the salvation of our God”)}
  \textbf{Gradual} (lesson chant performed after first reading) \\
  It is proper and corresponding text from text from Psalm 97, lines 2-4 (continues from the Introit) \\
  \textbf{Responsorial psalmody}: performed by a soloist alternating with a choir
  \begin{itemize}
    \item \textbf{monophonic} in texture.
    \item Pitch organization: \textbf{mode 5} (final - F, reciting tone - C)
    \item Melody constructed from
      \begin{itemize}
        \item multiple long \textbf{melismas}
        \item melodic formulas
      \end{itemize}
    \item Structure: 
    \begin{itemize}
      \item Response (opening by soloist and rest by choir). Material is \textbf{melismatic} (ornate and dozens of notes per syllable)
      \item Verse (marked V.) mostly soloist and choir enters at the end (entrance marked by *). Material is \textbf{melismatic}
    \end{itemize}
    \item Usually played as a musical highpoint of mass
  \end{itemize}
  \subsection{Types of Church Music}
  \subsubsection{Chants}
  \textbf{non-literate/unwritten} traditions where singers reconstruct melodies in performance using melodic formulas, performance conventions, and rules of Latin grammar.\\
  Important to note that this is neither memorized nor improvised. \\ \\
  Music notation was used for newly composed chant repertoire
  \begin{itemize}
    \item Sequences
    \item \textbf{Tropes}: textual or musical insertions into the body of the original chant and are usually sung by a solo voice. Used to elaborate on the theme of the day or explain something. \\
      Important to note that because this is adding to the original chant, tropes make the chant \textbf{proper}
  \end{itemize}
  \subsubsection{Organum}
  Earliest type of \textbf{polyphony} that consists of a 2-part texture 
  \begin{itemize}
    \item lower voice chant \textbf{(tenor)} sung by choir
    \item newly composed upper voice (\textbf{duplum}) sung by soloist
  \end{itemize}
  Typical organum practice during this time was \textbf{melismatic}
  \begin{itemize}
    \item slow, drone-like tenor
    \item fast, melismatic duplum
  \end{itemize}
  \textbf{Magnus Liber Organi} (“The Great Book of Organum”): collection of organum composed by \textbf{Master Leonin} and later revised by \textbf{Master Perotin}
  \begin{itemize}
    \item Above names were cited by \textbf{Anonymous IV}, a student (studied at \textbf{The School of Notre Dame}) from that era that named the composers and the works.
  \end{itemize}
  School of Notre-Dame brought a major innovation to music: \textbf{Rhythmic modes} system of rhythmic organization for organum for both composition and notation.
  \begin{itemize}
    \item notes are notated as either \textbf{long} or \textbf{breve} 
    \item Rhythmic patterns are grouped into \textbf{perfections} (units of 3)
  \end{itemize}
  Leonin's style, called \textbf{organum purum} ("pure organum") comprised of
  \begin{itemize}
    \item slow tenor, in free rhythm
    \item fast duplum in rhythmic modes
  \end{itemize}
  Perotin's style used \textbf{discants} (long melismas). In particular, when Perotin revised Magnus Liber, he replaced the organum purum with discants and composed new \textbf{discant clausulae} ("clauses"): long sections of discants that singers could substitute at will.
  \begin{itemize}
    \item fast tenor, in rhythmic modes
    \item fast duplum in rhythmic modes
  \end{itemize}
  \subsection{Leonin's and Perotin's setting of Viderunt omnes}
  \subsubsection{Leonin's Viderun omnes}
  Polyphony only seen in solo sections, choral sections are monophonic. \\
  The polyphonic portions can be split into 2 voices (tenor and duplum) and use organum purum style
  \subsubsection{Perotin's Viderun omnes}
  Consists of 4 voices that layer on top of each other, starting with the tenor (lowest)
  \begin{itemize}
    \item tenor
    \item duplum
    \item triplum
    \item quadruplum
  \end{itemize}
  Sections alternate between organum purum and discant clausulae. \\
  Upper voices share the same range and melodic ideas (\textbf{voice exchange}) \\ \\
  Vertical texture is a coincidence is a by product of the horizontal lines. Echo effect comes from the perfect consonances (4ths, 5ths, 8ves) \\
  During this time, 3rds and 6ths were seen as dissonant and thus were more hidden in the texture. Instead, 2nds and 7ths were more consonant and more prevalent in them music.
  \subsection{Secular Music and Types of Secular Music}
  Secular music was mostly seen in royal courts, festivals, parties.
  \subsubsection{Estampie}
  Popular dance at the time. 
  \begin{itemize}
    \item primary purpose of the music was to support the dancing so percussion instruments in particular were useful for counting steps.
    \item Monophonic tune was performed by an ensemble of loud instruments so that the music could be heard across the royal hall
    \item Structure consists of series of strains capped with open and then closed cadences
  \end{itemize}
  \subsubsection{Trouvère song}
  Monophonic secular song. Historically:
  \begin{itemize}
    \item Before ca. 1250: was used exclusively as courtly love poetry and songs for royalty and aristrocracy
    \item ca. 1250-1300: primary urban tradition of love poetry and songs where the royalty and \textbf{Puys} (brotherhood) sponsored performances
  \end{itemize}
  Genres include
  \begin{itemize}
    \item Chanson de jeste – a narrative epic
    \item Chanson courtoise – a song of courtly love
    \item Chanson de toile – a “woman’s song” (aka “spinning song”)
    \item Pastourelle – a pastoral
    \item Jeu-parti – a “mock debate”
  \end{itemize}
  \subsubsection{Motet}
  Most sophisticated "learned" secular genre that consists of 3 voices that are layered (tenor on bottom)
  \begin{itemize}
    \item Triplum (newly composed and based on French secular text)
    \item Duplum (newly composed and based on French secular text)
    \item Tenor (pre-existing untexted chant melisma)
  \end{itemize}
  \textbf{Adam de la Halle (d. ca. 1307)}: notable poet and composer who represented the last generation of trouvères. Was famous for both the poetic and musical aspects of trouvères (also famous for his motets). Normally only the poet is recognized and composer is not recognized.
  \subsection{Jeu de Robin et Marion (“play with songs”)}
  Trouvère song written by Adam de la Halle with a plot focuses on love story between Robin Hood and Maid Marion in a poetic/musical style. Song is organized into \textbf{refrains} (repeated with modifications) that make up a \textbf{rondeau} (round dance song). In particular, the structure is ABaabAB where lower cases are the refrains.
  \subsection{De ma dame vient/Dieus, comment porroie/Omnes}
  3-voice \textbf{Franconian motet} written by Adam de la Halle:
  \begin{itemize}
    \item 4 types of not durations that are grouped into perfections
    \item \textbf{polytextual}: duplum and triplum are set to different poems (representing the female lover and male lover that perform simultaneously but out of sync). Setting to different poems were used for comical effects. \\
  \end{itemize}
  Tenor part, which serves as a \textbf{cantus firmus} in particular is the opening melisma omnes from Gradual Viderunt omnes
  \begin{itemize}
    \item untexted and usually performed by an instrument, rather than voice
    \item slower than the upper 2 voices
  \end{itemize}
  Music style feature linear polyphony and the upper voices share the same range, with an emphasis on perfect consonances and accidental instances of dissonance.
  \subsection{Emergence of Polyphony}
  Hard to trace exact origins of polyphony but evidence suggests that early polyphonic music focused on amplifying monophonic chant, rather than serve as individual compositions. \\ \\
  Earliest recount of simultaneous sounding consonances are found in writings of \textbf{Hucbald} and \textbf{Regino} in the 9th century. The first writing of unambiguous polyphonic singing is in \textbf{Musica enchiradis} from the early 10th century. \\ \\
  Examples of polyphonic execution
  \begin{itemize}
    \item Wales: instrument execution was fast and lively 
    \item Britain: 2 part singing, 1 in a low register and the other singing something soothing to the ear at a higher range
  \end{itemize}
  Earliest polyphonic schools were centered around monasteries in France, England, Spain. It reached the School of Notre-Dame in the 12th/13th centuries and musicians from Paris were the first to notate both pitch and rhythm. \textbf{Anonymus IV} in his notes recounts several important figures during this time period. Most notable were
  \begin{itemize}
    \item \textbf{Master Leonius} known as best composer of organum
    \item \textbf{Master Perotin} best of composer of discant and introduced form of 4 voice writing (explained in previous section)
  \end{itemize}
  \section{At the Court of Philip the Good of Burgundy, ca. 1450}
  \subsection{The Duchy of Burgundy}
  One of the most important political, intellectual, and cultural centers of 15th century Europe.
  \begin{itemize}
    \item United large territories in France and the Low Countries
    \item played a key role in issues between European countries (100 year war, influence on several countries)
    \item Nominal capital was Dijon, but capital also moved between Brussels, Bruges, Lille, and Ghent
    \item \textbf{Phillip III (Philip the Good)} ruled between 1419-1467 and had an immense influence on Europe and expanded Dutch power. Philip the Good also supported scholars, writers, artists, musicians.
      \begin{itemize}
        \item \textbf{Philip's court chapel}: one of Europe's largest musical establishment that performed both secular and church music. Musicians at the chapel were well compensated.
        \item \textbf{Burgundian cathedral schools}: had the best musical and general education in Europe that basically guaranteed jobs for graduates. Musicians were so good that they were know was \textbf{Franco-Flemish} and were fought over by other European courts.
        \item \textbf{Philip's library}: one of Europe largest libraries that had manuscripts and musical manuscripts.
        \item \textbf{Feast of the Pheasant}: grand banquet hosted by Philip the Good where him and his knights publicly took the "Oath of the Pheasant"; a vow to launch a crusade to free Constantinople from the Turks.
          \begin{itemize}
            \item \textbf{L'homme arme (The Armed Man)}: monophonic chanson performed at the feast. Featured militant text with a melodic trumpet call to encourage Christians to arm themselves for battle. The tune ended up being a popular Mass cantus-firmus.
          \end{itemize}
      \end{itemize}
    \item \textbf{Guillaume de Machaut}: celebrated French poet and composer
      \begin{itemize}
        \item served variety of noble patrons across Europe
        \item associated with Cathedral of Notre-Dame of Rheims
        \item works were primary composed in form fixes
          \begin{itemize}
            \item \textbf{Ballade} - aabC
            \item \textbf{Rondeua} - ABaAabAB
            \item \textbf{Virelai} - AbbaA
          \end{itemize}
          These works typically consisted of 3 voices using the \textbf{ballade/cantilena} style
          \begin{itemize}
            \item Lively melody in the \textbf{cantus} (dominant voice, written first)
            \item \textbf{Tenor}: supports/cadences with the \textbf{cantus}
            \item \textbf{Contratenor}: middle voice (usually written last for textural filler)
          \end{itemize}
      \end{itemize}
  \end{itemize}
  \subsection{Rose liz printemps verdure}
  Written by Guillaume de Machaut in a 4-voice rondeau (ABaAabAB) where the tune is shared by the cantus and triplum.
  \begin{itemize}
    \item \textbf{Cantilena style}: only cantus is texted, all other voices are instrumental
    \item style is expanded in size (long refrain)
    \item Uses rhythmic and notational innovations of \textbf{Ars Nova} (duple and triple meter, syncopation, hemiola)
    \item Decorative melismas start each poetic line
  \end{itemize}
  The Rondeau's structure is also emphasized poetically by rhymes and musically by
  \begin{itemize}
    \item Hierarchy of cadences: strong at the end of the refrain (\textbf{double-leading-tone}) and weaker at half-points (\textbf{Phrygian})
    \item \textbf{Musical rhyme}: same music ends both halves of the refrain (with different cadences)
  \end{itemize}
  \subsection{Gilles Binchois (Gilles de Bins)}
  Composer, singere, and organist trained at a cathedral school. Worked in Philip the Good's court chapels until retirement and primarily composed 3-voice chansons (rondeaux and ballades)
  \subsubsection{Deuil angoisseux (ceremonial ballade (aa'bC))}
  \begin{itemize}
    \item Setting of the famous poetic lament by Christine de Pisane. 
    \item  Music is written in Cantilena style (tune in the cantus and supported by tenor/contratenor).
    \item Also uses musical rhyme (a' and C end with the same melisma; b and C start the same)
    \item Music is notably bright, major-ish harmonies
  \end{itemize}
  \subsection{The English Guise}
  Sweeter, more consonant music style preferred in English that traveled to Burgundy with the English armies during the 100-Year War. Burgundian musicians also it learned while working at English continental courts and by interacting with English composers in Burgundy. Characteristics include
  \begin{itemize}
    \item Strong preference for imperfect consonances (full triad, parallel 3rds/6ths, chains of 6/3 chords)
    \item Careful control of dissonance (prepare/resolve chords, but never on strong beats/important words)
    \item Syllabic text settings in a homophonic texture
    \item Burgundian composers merged guises with their own compositional material, resulting in a new international musical style: \textbf{Franco-Flemish polyphony}.
  \end{itemize}
  \subsubsection{John Dunstable}
  Known for his English guise
  \begin{itemize}
    \item Primarily sacred genres (Mass movements, motets)
    \item Preference for consonant 3 voice setting 
    \item Strongly consonant, triad-dominated textures (e.g. \textbf{Quam pulchra es})
  \end{itemize}
  \subsubsection{Quam pulchra es}
  3-voice motet (short setting of Latin sacred text)
  \begin{itemize}
    \item Freely composed setting (no borrowed chant) 
    \item Text from \textbf{The Song of Songs} is set syllabically (musical structural, rhythmic patterns, cadence placement guided by the poetry)
    \item Homophonic texture filled with triads and 6/3 chords
  \end{itemize}
  \subsubsection{Johannes Tinctoris}
  Burgundian theorist and composer who claimed in his treatise \textbf{Liber de arte contrapucti} that Franco-Flemish music has reached a state of perfection and that everything written before it was not worth listening to.
  \subsubsection{Martin le Franc}
  Burgundian court poet who also claimed that early French composers' music pales to that of Burgundian composers who have learned to use pleasant consonant sound.
  \subsubsection{Guillaume Du Fay}
  Had an illustrious international career and was known to be the greatest Burgundian composer of his generation.
  \begin{itemize}
    \item Honorary member of Philip of Burgundy's court chapel
    \item Had many European noble patrons
    \item \textbf{Nuper rosarum flores}: Du Fay's grant motet for the dedication ceremony (dedication of Cathedral of Santa Maria del Fiore)
    \item Standardized the first type of polyphonic Mass Ordinary (\textbf{cyclical cantus-firmus Mass})
      \begin{itemize}
        \item Signature genre of Franco-Flemish composers
        \item consists of 5 movements
        \item Same borrowed melody in the tenor for all movements (\textbf{cantus-firmus}). This can be from a chant or from a secular song.
      \end{itemize}
  \end{itemize}
  \subsubsection{Nuper rosarum flores}
  Consists of 4 voices
  \begin{itemize}
    \item 2 tenors
    \item motetus
    \item triplum
  \end{itemize}
  Borrows the chant \textbf{Terribilis est locus iste}, an introit of the Mass for the dedication of the church \\ \\
  Some important number symbolism to note
  \begin{itemize}
    \item 7 ("Virgin Mary"): lines per stanza, syllables per line
    \item 4 ("temple"): number of poetic stanzas, voices, sections
    \item 14 ("days of Temple dedication feast"): number of pitches in chant fragment
    \item 28 (7 * 4, 14 *2): total number of poetic lines; total number of pitches in 2 tenors of each section; total number of \textbf{tempora} (time units) in each section
  \end{itemize}
  Chant fragment is written out once and is completed by a \textbf{canon} (instructions on how to perform it). \\ \\
  Is an \textbf{isorhythmic} motet that features repeated \textbf{colors} (repeated pitch pattern) and \textbf{Talea} (repeated rhythmic pattern)
  \subsubsection{Missa L'homme arme}
  Cantus-firmus Mass by Guillaume Du Fay (first known example of L'homme armee Mass)
  \begin{itemize}
    \item Borrowed melody appears at least once in the tenor of all 5 movements
    \item Consists of 4 voices (SATB quartet) that do not share range
    \item Reduced textures, with sections beginning with a \textbf{duet}
    \item \textbf{Motto technique}: all movements begin with a variant of the same duet
    \item Melodic/rhythmic gestures of L'homme arme appear in the other voices
  \end{itemize}
  \textbf{Gloria} cantus-firmus appears once per section. Each section starts with a duet then continues in a 4-part texture 
  \begin{itemize}
    \item 1st statement of cantus-firmus in long notes
    \item 2nd statement is 1/3 faster; creating a hemiola
  \end{itemize}
  \textbf{Agnus Del}: 3 distince sections, each treats the cantus-firmus differently
  \begin{itemize}
    \item Agnus-I: starts with a duet, then 1 complete statement in the tenor
    \item Agnus-II: faster, 3-voice section (no tenor); opening cantus-altus duet in imitation, then a variant of the L'homme arme tune fast in the bassus
    \item Agnus-III: returns to slower pace, 4 voices. 2 statements of cantus firmus: slow in retrograde, then in its original form but melodically ornamented and twice as fast.
  \end{itemize}

  \subsection{Messe de Nostre-Dame}
  Written by Guillaume de Machaut and was the first known polyphonic setting of the Mass Ordinary.
  \begin{itemize}
    \item 4 voices sing continuously (no monophonic sections) 
    \item Characteristic descending motif in all movements (a unifying device)
    \item \textbf{Gloria, Credo}: freely composed (no chant)
    \item \textbf{Kyrie, Sanctus, Agnus Del, Ite Missa Est, Amen of Credo}: (texts based on local Rheims-area chants that are used in \textbf{isorhythmic procedure} in the tenor
  \end{itemize}
  \section{In the Printshops of Ottaviano Petrucci and Pierre Attaingnant ca. 1500-1550}
  \textbf{Economic changes with the printing press}: \\
  Invention of printing made access to music easier, making it as a possible business endeavor since people could afford to purchase music. Furthermore, music could travel faster across territories since the composer didn't need to be present to deliver the music. \\ \\
  \textbf{Changes to musicians} \\
  Composers could sell music to publishers and thus could support themselves rather than relying on patronage. Furthermore, musicians could gain access to a broader range of up-to-date music.
  \subsection{Ottaviano Petrucci}
  First publisher to print polyphonic music that used \textbf{triple-impressions} a slow process but produced precise and visually pleasing scores. He ended up creating a monopoly of music publishing in Venice ca. 1500-1520, creating 67 volumes of polyphonic music from both Franco-Flemish and Italian pieces. The volumes were very popular and were constantly reprinted. \\ \\
  \textbf{Harmonice Musices Odhecaton} (100 Polyphonic Songs") was his first publication, a collection of Franco-Flemish chansons. It was so popiular that Petrucci printed 2 more volumes (\textbf{Canti B} and \textbf{Canti C}) \\ \\
  Next set of publication was Masses written by \textbf{Josquin des Prez}
  \subsection{Josquin des Prez (ca. 1450-1521)}
  Franco-Flemish known as the best composer of his time. He trained at the Cambrai cathedral school and was well versed in the material from his predecessors (had small tributes to other composers in his music). Because os Josquin's notoriety, other composers would often put his name on their music to ensure that it sold. Josquin eventually ended up signing his music to prevent copyright. \\ \\
  His Masses consist both of traditional Franco-Flemish and new types:
  \begin{itemize}
    \item Cantus-firmus Mass
    \item Plainsong Mass: each movement's cantus-firmus is a separate chant
    \item Paraphrase Mass: variant of borrowed tune is in all voices via \textbf{imitation}
    \item Imitation or parody Mass: multiply voices of a polyphonic model are used.
  \end{itemize}
  \subsubsection{Missa L'homme arme super voces musicales}
  First published Mass Ordinary in print. Opens Petrucci's first volume of Josquin Masses. It is set to the traditional L'homme arme tune using
  \begin{itemize}
    \item voces musicales: solmization syllables: ut, re, mi, fa, sol, la
    \item cantus-firmus transposed a step upward in each movement
    \item contains several sophisticated craftsmanship (references to predecessors) that only clever audiences would understand
    \item features multiple \textbf{mensuration canons} in each movement (multiples singers the same melody at different speeds in different meters (mensurations)). However, the singers are given a set of difficult instructions (canons) from the same line of music
    \item \textbf{Kyrie} movement: L'homme arme tune is divided into 3 parts. In each section, the tune is set in a 2-voice mensuration canon where the cantus-firmus in the tenor is the slower the answering voice.
  \end{itemize}
  \subsubsection{Missa Pange Lingua}
  Possibly Josquin's last Mass and is a paraphrase Mass based on \textbf{Pange lingua gloriosi} that consists of 4 voices that are consonant and more textually fluid \\ \\
  \textbf{Paraphrase Mass}: borrowed melody that is melodically varied in all 5 movements and is shared between all 4 voices using \textbf{imitation}. \\ \\
  In the \textbf{Kyrie} movement:
  \begin{itemize}
    \item Each phrase of the hymn is a subject for a \textbf{point of imitation}, followed by a drive to the cadence, which ends a section
    \item each section uses \textbf{paired imitation} (pair of voices in imitation is imitated by another pair of voices)
  \end{itemize}
  \subsubsection{Ave Maria... virgo serena}
  Josquin's earliest motet and opens Petrucci's first volume of Josquin motets. Was reprinted multiple times (very popular).
  \begin{itemize}
    \item Text is a conflation of 3 popular Marian prayers and hymns
    \item Music was a chant that set the opening sequence test of the Feat of Annunciation (variant from Milan)
    \item Known for its transparent and sensitive text setting (each line of text is set to an individual phrase).
    \item Also features cadences denoted ends of sections that align with the poetry \text{syntactic immitation}
    \item The chant melody is paraphrased throughout the motet but is NOT a cantus firmus (focused on all 4 voices rather than single voice)
  \end{itemize}
  \subsection{Rise of Native Italian Music}
  Although Oltremontani composers dominated the music markets, Italian secular began to rise in popularity in the early 1500. In particular, Petrucci printed 11 collections of \textbf{frottola}, an Italian strophic dance-like song with chordal accompaniment
  \begin{itemize}
    \item Court entertainment: semi-popular texts are set in a popular style written and composed by professionals.
    \item Developed from a unique Italian tradition of improvising lyric poetry sung to a strumming accompaniment.
    \item Genre was cultivated at the Gonzaga court, duchy of Mantua
    \item Isabella Gonzaga: musician and patron of the arts and employed \textbf{Marchetto cara}, a frottola composer.
    \item Marchetto cara's compositions had an impact on Frottola music - Petrucci printed a set of rules for frottola compositions.
  \end{itemize}
  \subsubsection{lo non compro piu speranza}
  Popular frottola written by Marchetto Cara. Only had a Cantus and Tenor; performer can play the bass line as written or using improvised root-position triads.
  \begin{itemize}
    \item Duple and triple rhythms alternate, creating a hemiola effect
    \item Syllabic text setting with melisma on the penultimate syllable
    \item Poetic/musical structure audible with repetitions and clear cadences
    \item Although it competed with the Franco-Flemish chanson, frottola eventually died out in popularity.
  \end{itemize}
  \subsection{Pierre Attaingnant (1494-1552)}
  The next big Parisian publisher who had a monopoly on printing polyphonic music during his time. \\ \\
  Printed music that consisted of a staff, notes, and text. Alignment was very difficult to produce, so the music was less visually pleasing. However, it was cheap and fast to produce. \\ \\
  Core repertoire he printed was \textbf{chansons nouvelle} (\textbf{Parisian chansons}), a new French genre distinct from the Franco-Flemish chanson.
  \begin{itemize}
    \item Secular song based on humorous courtly love poetry.
    \item Prominent melody: polyphonically animated homophonic texture.
  \end{itemize}
  Pierre Attaingnant also published his own song arrangements and collections of instrumental music in \textbf{tablature} (instrument-specific notation that sometimes accompanied texts (frotolla)) \\ \\
  People often bought partbooks of (untexted) vocal polyphony and \textbf{intabulations} (instrumental arrangements of vocal music) to play on their on instruments. Dance music was also popular and were printed in partbooks with instrumentation. \\ \\
  Pierre Attaingnant's dance music came in \textbf{danseries}: multiple volumes of dance suits, usually sets of 2 that consisted of
  \begin{itemize}
    \item slower duple-time dance
    \item faster one in triple-time
  \end{itemize}
  \subsubsection{Tant que vivray}
  Chanson nouvelle by Claudin de Seermisy
  \begin{itemize}
    \item Music set to a poem by Clement Marot
    \item Syllabic text setting: attention to \textbf{prosody} (accents on strong beats; long notes on mid-line caesuras; cadences on punctuation marks)
    \item Musical structure: 2 sections:
      \begin{itemize}
        \item A (repeated): slow, longer phrases in homo-rhythm
        \item B: fast, repetitive shorter phrases on an animated texture
      \end{itemize}
    \item uses \textbf{musical rhyme}: both A and B end on the same cadence
  \end{itemize}
  \section{The Year 1600: Medici's Florence and Elizabeth's London}
  \subsection{Polyphonic Madrigal}
  Leading genre of secular music during late Renaissance that was a consequence of the \textbf{Petrachan movement}: led by \textbf{Pietro Bambo} who aimed at raising the Tuscan dialect of Florence.
  \begin{itemize}
    \item Poetic style: borrowed from \textbf{trecento madrigal} (forme fixe)
    \item Musical style: fusion of frottola and Franco-Flemish motet with high-quality Italian poetry and homophonic and imitative textures. Furthermore, each poetic line receives individual treatment.
    \item Singers were each assigned one part (motet style)
  \end{itemize}
  Early madrigals were composed by Franco-Flemish composers (e.g. Philippe Verdelot and Jacques Arcadelt) but were eventually composed by Italian composers ca. 1600 (e.g. Carlo Gesualdo and Claudio Moteverdi)
  \subsection{Luca Marenzlo}
  One of the greatest Italian born madrigalists who had international popularity. His music was particular difficult to perform and was intended for trained professionals.
  \subsubsection{Solo e pensoso}
  From Marenzlo's 9th book of madrigals
  \begin{itemize}
    \item 5-voice polyphonic madrigal set to a sonnet by Petrarch
    \item Sonnet is split into 2 sections
    \item Each poetic line receives individual musical treatment/phrase
    \item Uses \textbf{Chromaticism}: pitches/chords outside the mode, resulting in a chromatic line
    \item Uses \textbf{word painting} (madrigalism): imitating meaning of a word with a musical equivalent (e.g. melodic contour)
  \end{itemize}
  \subsubsection{Concerti delle donne}
  Professional female vocal ensembles of late Renaissance Italy that performed the madrigals. First established ca. 1580.
  \subsection{Monody}
  New and improved type of solo madrigal (based on ancient Greek dramas). Solo vocal line (or multiple lines in unison) was set to a text with virtuosic embellishments and was accompanied by \textbf{basso continuo} (notated in \textbf{figured bass}). \\ \\
  Was argued by \textbf{Girolamo Mei (1519-1594)} who claimed that they should return to Ancient Greek musical practices. Theorized that
  \begin{itemize}
    \item Music should move people by imitating speech through pitch and rhythm (single line)
    \item Part singing is never mentioned in ancient Greek description, hence it doesn't exist
    \item Modern polyphony fails to moving people. 
  \end{itemize}
  In practice, monody was
  \begin{itemize}
    \item Created by singers who composed, rather than composed by professional composers
    \item Concentrated on melody highlights of singer's virtuosity and text
  \end{itemize}
  \subsubsection{The Floretine Camerata}
  A group of intellectuals who gathered informally to discuss the latest theories on Ancient Greek Music and its use.
  \subsection{Giulio Caccini}
  Composer and singer at the Medici court.
  \subsubsection{Le Nuove Muisiche}
  Collection of solo madrigals by Caccini. All pieces are monodies printed with figured bass. Includes a lengthy preface that sets rules for the text setting and performance practice. \\ \\
  \textbf{Amarilli, mia bella} one of the most famous solo madrigal set to Alessandro Guarini. Syllabic text setting with ornaments on accented syllables.
  \subsection{Staged Music in Medici Florence}
  Lavish court entertainment, e.g. \textbf{intermedi}, for the elite and invited audience.
  \begin{itemize}
    \item Staged musical entertainment between acts of spoken play
    \item Plots are unrelated to the play and emphasize of spectacle or is tied to the social occasion for which they are staged.
  \end{itemize}
  Most famous Intermedi: \textbf{La pellegrina}: a set of 6 intermedi created for a Medici wedding based on 6 different plots:
  \begin{itemize}
    \item Harmony of the Spheres: power of music
    \item Singing Contest between Pierides and Muses: power of music
    \item Apollo Slays the Monster at Delphi: power of music
    \item The Golden Age is Foretold: glorification of the royal couple
    \item Arion Saved by the Dolphin: power of music
    \item Jove's Gift to mortals of Rhythm and Harmony: power of music, glorification of the royal couple
  \end{itemize}
  \subsubsection{Peri, Euridice (1600)}
  First surviving opera; a play (\textbf{libretto}) full set to music. \\ \\
  \textbf{Scene with the Messenger (Act 2)}: a pivotal dramatic moment in the story. Is set in \textbf{recitative}: declamatory vocal lines over sustained continuo chords.
  \begin{itemize}
    \item Voices move freely between dissonance and consonances, imitating speech
    \item Basso continuo chords change at the main accent of each poetic line.
  \end{itemize}
  \subsection{Elizabethan England}
  Madrigals were introduced to England in the book \textbf{Musica Transalpina}, an anthology of popular Italian polyphonic madrigals. This book ended up being very popular and was republished several times and started an interest for English madrigals at Elizabeth courts. \\ \\
  Genres of secular songs at the time
  \begin{itemize}
    \item \textbf{English madrigal}: Italian style polyphonic madrigal in English
    \item \textbf{Consort song}: solo song accompanied by a consort of viols
    \item \textbf{Lute ayre}: solo song with lute accompaniment
  \end{itemize}
  \subsubsection{John Dowland (1562-1626)}
  Elizabethan lutenist and composer who composed multiple vocal and instrumental genres for court chamber
  \subsubsection{Flow my tears}
  Lute ayre written by Dowland set to his own text in repeated melodic strains (AABBCC). Structure of a \textbf{pravane} (originated as a dance style in Italy). Mournful text reflected by descending melodic contour.
  \subsubsection{The virginal}
  Popular keyboard instrument during Elizabeth's court. Was compact and used to accompany vocal music and specific keyboard repertoire. \textbf{Fitzwilliam Virginal Book}: collection of Elizabethan music for the virginal
  \subsubsection{Pavana Lacrimae}
  Written by William byrd and is a variation of Dowland's Flow my tears. Consists of sections variations (first introduced as an intabulation then as a variation).
  \subsubsection{Consort music}
  Music arranged for a consort of viols, including transcriptions and intabulations of vocal music or variations on pre-existing melodies:
  \subsection{Elizabethan Theater Music}
  Cross-pollination of court and town music, even in Shakespeare plays, designed for the common people:
  \begin{itemize}
    \item Opening and closing music: fanfares
    \item Embedded songs: popular tunes of diverse origins borrowed from texts performed by main characters or musicians of the play.
    \item Music is based on what is popular at the time, is not written for the play usually
  \end{itemize}
  \begin{itemize}
    \item Sacred
    \item Secular
  \end{itemize}
 \end{document}
